Reconstruction of Large Scale Structure by Nonlinear Displacement Fields

### 2016-09-24 ms_disp1/ms_disp1.tex
haoran

\begin{abstract}
We investigate the ability of using nonlinear displacement field to reconstruct 
the primordial linear perturbations of the large scale structure. According to 
the linear Lagrangian perturbation theory, the first order primordial linear 
density field is given by the negative divergence of the displacement field. By 
running $N$-body simulations, we show that this reconstruction algorithm recovers 
great amount of nonlinear information in the early universe. In reality, the 
displacement fields can be solved from only the final stage of the density 
fields. This has potential to reconstruct baryonic acoustic oscillation (BAO) 
from current and future large scale structure surveys.

@@

\section{Introduction}\label{sec.intro}
Our universe starts from primordial Gaussian perturbations at a very early stage, 
and from those fluctuations, the gravitational instability drives the formation of 
the large scale structure (LSS) distribution of matter. These structures grow 
linearly until the perturbations are large enough such that first order 
perturbation theories are unable to analytically describe the LSS distributions. 
Inversely, the final stage nonlinear LSS distribution contains higher order 
statistics, and thus makes it more challenging to interpret the basic cosmological 
parameters. This leads to various attempts to recover earlier stages of LSS, in 
which statistics are closer to Gaussian. Because Gaussian fields can be adequately 
described by two-point statistics, ideally after some recovery algorithms, more 
information can be extracted, more straightforwardly, by power spectra or two-point 
correlation functions.

These algorithms include nonlinear Wiener filters in wavelet space, distribution 
function transforms etc. Standard BAO reconstruction algorithms smooths the 
nonlinear density field on linear scale ($\sim$10 Mpc$/h$) and reverse the large 
scale bulk flows by a negative Zel'dovich linear displacement. Here we propose a 
new reconstruction method that uses the nonlinear displacement field to recover the 
primordial density field. In the linear Lagrangian perturbation theory, the 
negative divergence of the the displacement field respect to Lagrangian coordinates 
gives the linear density field. Of course, the full displacement field is non-
observable, as it requires the initial distributions of matter, however there are 
many techniques to estimate the nonlinear displacement field from a final 
distribution of matter. For example, when a homogeneous initial matter distribution 
is assumed, there is a unique solution of curl-less displacement field to relate 
the initial and final distributions without shell-crossing. This solution can be 
solved by a metric transformation equation \citep{1995ApJS..100..269P,1998ApJS..
115...19P}. In 1-dimensional case, the solution simplifies to a reordering of 
matter elements by Eulerian positions, and \cite{2016arXiv160907041Z} studies this 
case. The homogeneous and curl-less assumptions are generally valid in cosmology 
[cite or proof]. Shell-crossing does exist, and this effect can be quantified by 
$N$-body simulations.

In this paper we track the initial and final distributions of particles in LSS $N$-
body simulations, and use the displacement field to reconstruct the initial linear 
density field. The rest of this paper is structured as follows. In section 
\ref{sec.method} we describe the simulation and reconstruction algorithm. In 
section \ref{sec.results} we show the results of the reconstruction. Discussion and 
conclusion are in section \ref{sec.discussion}.

@@

\section{Method}\label{sec.method}

\subsection{Simulation}
We use the public cosmological simulation code {\tt CUBE}. Cosmological parameters 
are in accordance with Planck. Initial conditions are generated at redshift $z=50$ 
using Zel'dovich approximation. $N_p=512^3$ $N$-body particles are evolved via 
their mutual gravitational interactions to $z=0$, in a periodic box with $L=300$ 
Mpc$/h$ per side. The code is set to use standard a particle-mesh (PM) algorithm 
\cite{1988csup.book.....H} on a two-level mesh grids (details see \cite{2013MNRAS.
436..540H}) and cloud-in-cell (CIC) is used in particle interpolations in force 
calculation and obtaining the density field $\rho({\bs x})$ in Eulerian coordinates 
${\bs x}$ at late stages. We use density contrast $\delta\equiv\rho/\lb\rho\rb-1$ 
to describe the density fluctuations. The primordial linear density field $
\delta_L$ is given by the initial stage and scaled to $z=0$ by the linear growth 
factor.

Two-point statistics of the density fields are quantified by the cross power 
spectrum $P_{ij}(k)\equiv(2\pi)^{-3}\langle|\delta_i(k)||\delta_j(k)|\rangle$, 
where subscripts $i,j$ may refer to linear, nonlinear, or reconstructed density 
fields. When $i=j$ it reduces to the auto power spectrum $P_{ii}(k)$ or $P(k)$. We 
usually plot the dimensionless power spectrum $\Delta^2(k)\equiv k^3P(k)/2\pi^2$. 
The blue solid and black dashed curves in Fig.\ref{fig.1} show the power spectra of 
$\delta$ and $\delta_L$. Their difference shows the nonlinear evolution of LSS on 
small scales. Their cross power drops to a very low value, indicating a loss of 
linear information in the nonlinear power spectrum.

\subsection{Reconstruction}
In the simulation, we uses particle-ID (PID) to record the initial location ${\bs 
q}$ of particles, and the information is tracked until the $z=0$ and we can get the 
Lagrangian displacement vector ${\bs \Psi}\equiv{\bs x}-{\bs q}$ for every 
particle. Then these vectors are interpolated onto the initial Lagrangian 
coordinates ${\bs q}$ of particles and we get the displacement field ${\bs \Psi}
({\bs q})$.
The raw reconstructed density field is given by the differential motion of matter 
elements,
\begin{equation}
    \delta_R=-\nabla\cdot{\bs \Psi}({\bs q}).
\end{equation}
In Fig.\ref{fig.1} we show the power spectrum of $\delta_R$ and its cross power with $\delta_L$. Despite of a lowered power of $\delta_R$ compared to $\delta$, it has a much higher cross power with $\delta_L$, up to a relatively smaller scale (higher $k$).

>>>>

\begin{abstract}
We investigate the ability of using nonlinear displacement field to reconstruct 
the primordial linear perturbations of the large scale structure. According to 
the linear Lagrangian perturbation theory, the first order primordial linear 
density field is given by the negative divergence of the displacement field. By 
running $N$-body simulations, we show that this reconstruction algorithm recovers 
great amount of nonlinear information in the early universe. In reality, the 
displacement fields can be solved from only the final stage of the density 
fields. This demonstrates the potential to reconstruct baryonic acoustic oscillation (BAO) 
from current and future large scale structure surveys.

@@

\section{Introduction}\label{sec.intro}
Our universe starts from primordial Gaussian perturbations at a very early stage, 
and from those fluctuations, the gravitational instability drives the formation of 
the large scale structure (LSS) distribution of matter [cite LSS]. These structures grow 
linearly until the perturbations are large enough so that the first order 
perturbation theories are unable to analytically describe the LSS distributions [cite PT]. 
As a result, the final nonlinear LSS distribution contains higher order 
statistics, and thus makes it more challenging to be interpreted into basic
cosmological parameters [cite constraint]. For example, the baryonic acoustic oscillation (BAO)
scale can be used as a ``standard ruler'' to constrain the cosmic expansion history
and thus probes the dark energy properties [cite BAO], and nonlinear evolution smears the BAO
features [cite smear BAO]. These leads to various attempts to recover earlier stages of LSS, in 
which statistics are closer to Gaussian [cite Gaussianize, Reco examples].
Because Gaussian fields can be adequately described by two-point statistics,
ideally after some recovery algorithms, more information can be extracted,
more straightforwardly, by power spectra or two-point correlation functions [can cite friends].

These algorithms include nonlinear Wiener filters in wavelet space [cite myself], distribution 
function transforms etc [cite Pengjie, etc.]. Standard BAO reconstruction algorithms smooth the 
nonlinear density field on linear scale ($\sim$10 Mpc$/h$) and reverse the large 
scale bulk flows by a negative Zel'dovich linear displacement [cite Eisenstein]. Here we propose a 
new reconstruction method that uses the nonlinear displacement field to recover the 
primordial density field. In the linear Lagrangian perturbation theory, the 
negative divergence of the the displacement field respect to Lagrangian coordinates 
gives the linear density field [cite Jeong]. Of course, the full displacement field is non-
observable, as it requires the initial distributions of matter, however there are 
many techniques to estimate the nonlinear displacement field from a final 
distribution of matter. For example, when a homogeneous initial matter distribution 
is assumed, there is a unique solution of curl-less displacement field to relate 
the initial and final distributions without shell-crossing. This solution can be 
solved by a metric transformation equation \citep{1995ApJS..100..269P,1998ApJS..115...19P}. In 1-dimensional case, the solution simplifies to a reordering of 
matter elements by Eulerian positions, and \cite{2016arXiv160907041Z} studies this 
case. The homogeneous and curl-less assumptions are generally valid in cosmology 
[cite or proof]. Shell-crossing does exist, and this effect can be quantified by 
$N$-body simulations.

In this paper we track the initial and final distributions of particles in LSS $N$-
body simulations, and use the displacement field to reconstruct the initial linear 
density field. In section 
\ref{sec.method} we describe the simulation and reconstruction algorithm. In 
section \ref{sec.results} we show the results of the reconstruction. Discussion and 
conclusion are in section \ref{sec.discussion}.

@@

\section{Method}\label{sec.method}
We show the LSS simulation and displacement field setups in section \ref{ss.sim},
and the reconstruction processes are presented in section \ref{ss.reco}.

\subsection{Simulation}\label{ss.sim}
We use the open source cosmological simulation code {\tt CUBE}\footnote{github link of cafcube}
[cite Yu et al. in prep].
Cosmological parameters 
are [TBD before submission]. Initial conditions are generated at redshift $z=50$ 
using Zel'dovich approximation. $N_p=512^3$ $N$-body particles are evolved via 
their mutual gravitational interactions to $z=0$, in a periodic box with $L=300$ 
Mpc$/h$ per side. The code is set to use standard a particle-mesh (PM) algorithm 
\cite{1988csup.book.....H} on a two-level mesh grids
(details see \cite{2013MNRAS.436..540H}) and cloud-in-cell
(CIC) is used in particle interpolations in force 
calculation and obtaining the density field $\rho({\bs x})$ in Eulerian coordinates 
${\bs x}$ at late stages. We use density contrast $\delta\equiv\rho/\lb\rho\rb-1$ 
to describe the density fluctuations. The primordial linear density field $
\delta_L$ is given by the initial stage and scaled to $z=0$ by the linear growth 
factor.

Two-point statistics of the density fields are quantified by the cross power 
spectrum $P_{ij}(k)\equiv(2\pi)^{-3}\langle|\delta_i(k)||\delta_j(k)|\rangle$, 
where subscripts $i,j$ may refer to linear ($L$), nonlinear ($N$), or reconstructed ($R$) density 
fields. When $i=j$ it reduces to the auto power spectrum $P_{ii}(k)$ or $P(k)$. We 
usually plot the dimensionless power spectrum $\Delta^2(k)\equiv k^3P(k)/2\pi^2$. 
The blue solid and black dashed curves in Fig.\ref{fig.1} show the power spectra of 
$\delta$ and $\delta_L$. Their difference shows the nonlinear evolution of LSS on 
small scales. Their cross power drops to a very low value, indicating a loss of 
linear information in the nonlinear power spectrum.

\subsection{Reconstruction}\label{ss.reco}
In the simulation, we use particle-ID (PID) to record the initial (Lagrangian) location ${\bs 
q}$ of particles, and the information is tracked until the $z=0$ and we can get the 
Lagrangian displacement vector ${\bs \Psi}\equiv{\bs x}-{\bs q}$ for every 
particle. Then these vectors are interpolated onto the initial Lagrangian 
coordinates ${\bs q}$ of particles and we get the displacement field ${\bs \Psi}
({\bs q})$.
The raw reconstructed density field is given by the differential motion of matter 
elements,
\begin{equation}
    \delta_R=-\nabla\cdot{\bs \Psi}({\bs q}).
\end{equation}
Because the reconstruction processes are implemented on Lagrangian coordinates,
$\delta_R$ takes the coordinates of $\bs q$ instead of $\bs x$.
We just write $\bs q$'s Fourier wave number
$k_q$ as $k$ to simplify the expression.
In Fig.\ref{fig.1} we show the power spectrum of $\delta_R$ and its cross power with $\delta_L$. Despite of a lowered power of $\delta_R$ compared to $\delta$, it has a much higher cross power with $\delta_L$, up to a relatively smaller scale (higher $k$).
